---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# taxalight :zap: :zap:

<!-- badges: start -->
[![R build status](https://github.com/cboettig/taxalight/workflows/R-CMD-check/badge.svg)](https://github.com/cboettig/taxalight/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/taxalight)](https://CRAN.R-project.org/package=taxalight)
<!-- badges: end -->

`taxalight` provides a lightweight, lightning fast query for resolving taxonomic identifiers to taxonomic names, and vice versa, by using a Lightning Memory Mapped Database backend.


## Installation

You can install the released version of taxalight from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("taxalight")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cboettig/taxalight")
```

## Quickstart

`taxalight` needs to first download and import the provider naming databases.  This can take a while, but needs to only be done once.

```{r example}
library(taxalight)
tl_create("itis")
```

Now we can look up species by names, IDs, or both at once:

```{r}
tl("Homo sapiens")
```
```{r}
id <- c("ITIS:180092", "ITIS:179913", "Dendrocygna autumnalis")
tl(id)
```


For convenience, we can request just the name or id as a character vector (paralleling functionality in `taxize`):


```{r}
get_ids("Homo sapiens")
```


```{r}
get_names("ITIS:179913")
```

## Benchmarks

```{r}
library(bench)

```


```{r}
tl_create("col")
```

```{r}
duckdb <- DBI::dbConnect(duckdb::duckdb(), dbdir = tempfile())
taxadb::td_create("col", schema="dwc", db = duckdb)
```




```{r}
rsqlite <- DBI::dbConnect(RSQLite::SQLite(), dbdir = tempfile())
taxadb::td_create("col", schema="dwc", db = rsqlite) 
```
```{r}
ids <- c("COL:35517330", "COL:35517332", "COL:35517329", "COL:35517325")
bench::bench_time(
  
  taxadb::filter_id(ids, "col", db = rsqlite)
  
)
```


```{r}
bench::bench_time(
  
  taxadb::filter_id(ids, "col", db = duckdb)
  
)
```

```{r}
bench::bench_time(
  
  taxalight::tl(ids, "col")
  
)
```

## A provenance-backed data import

Under the hood, `taxalight` consumes a [DCAT2/PROV-O based description](https://raw.githubusercontent.com/boettiger-lab/taxadb-cache/master/prov.json) of the data provenance which generates the standard-format tables imported by `taxalight` (and `taxadb`) from the original data published by the naming providers.  All data and scripts are identified by content-based identifiers, which can be resolved by <https://hash-archive.org> or the R package, `contentid`.  This provides several benefits over resolving data from a URL source:

1. We have cryptographic certainty that we get the expected bytes every time
1. We can automatically cache and reference a local copy.  If the hash matches the requested identifier, then we don't even need to check eTags or other indications that the version we have already is the right one.
1. By registering multiple sources, the data can remain accessible even if one link rots away.  

Input data and scripts for transforming the data into the desired format are similarly archived and referenced by content identifiers in the provenance trace.  




