---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# taxalight :zap: :zap:

<!-- badges: start -->
[![R build status](https://github.com/cboettig/taxalight/workflows/R-CMD-check/badge.svg)](https://github.com/cboettig/taxalight/actions)
<!-- badges: end -->

`taxalight` provides a lightweight, lightning fast query for resolving taxonomic identifiers to taxonomic names, and vice versa, by using a Lightning Memory Mapped Database backend.


## Installation

You can install the released version of taxalight from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("taxalight")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cboettig/taxalight")
```

## Example

`taxalight` needs to first download and import the provider naming databases.  This can take a while, but needs to only be done once.

```{r example}
library(taxalight)
tl_create("itis")
```

Now we can look up species by names, IDs, or both at once:

```{r}
tl("Homo sapiens")
```

